---
title: "Agri-food chain employment"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

```

## Acquire data

The first step is to acquire the raw data.

```{r raw-data, warning=FALSE}

library(fsp)

lvls <- c("Agriculture",
          "Fishing",
          "Manufacturing",
          "Wholesale",
          "Retail",
          "Catering")

ag_gb <- fsp::compile_ag_workforce()
jobs03 <- fsp::get_jobs03()

food <- fsp::jobs03_sectors() |> 
  dplyr::left_join(jobs03, by = dplyr::join_by(sic_division)) |> 
  dplyr::group_by(sector, date) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::bind_rows(ag_gb) |> 
  dplyr::mutate(sector = forcats::fct(sector, levels = lvls))


#added total industries comparison
food_auk<- jobs03_AUK_sectors() |>
    dplyr::left_join(jobs03, by = dplyr::join_by(sic_division)) |> 
  dplyr::group_by(sector, date) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::bind_rows(ag_gb) #Q4 2023 figures not released?

food |> 
  dplyr::arrange(desc(date)) |> 
  head() |> 
  knitr::kable()

```

## Wrangle

Since the agricultural workforce data is annual, and the JOBS03 data is
quarterly, the latest year may not include figures for agriculture. Lets get
some yearly averages, and we will calculate some percentage changes in case they
come in useful.


```{r}

fsp_food <- food |> 
  dplyr::mutate(year = lubridate::year(date)) |> 
  dplyr::group_by(sector, year) |> 
  dplyr::summarise(value = mean(value)) |> 
  dplyr::group_by(sector) |> 
  dplyr::mutate(yoy_pct_change = scales::label_percent(accuracy = 0.1) (value/dplyr::lag(value) - 1)) |> 
  dplyr::group_by(year) |> 
  dplyr::mutate(pct_of_tot = scales::label_percent(accuracy = 0.1) (value/sum(value))) 

#latest_year<-max(fsp_food$year) #potentially parameter for the latest year

fsp_food |> 
  dplyr::filter(year == 2022) |>
  knitr::kable(digits = 2)


```


## Prepare for publication

Now we have got the numbers, its just a case of wrangling them into a table in a
suitable form for converting to govspeak and publication. This code feels clunky
and inelegant, but it works.


```{r govspeak, results='markup'}

x <- fsp_food |> 
dplyr::mutate(total = "Total Agri-Food",
              food_chain = dplyr::case_when(sector != "Agriculture" ~ "Total Food"),
              ag_fish = dplyr::case_when(sector %in% c("Agriculture", "Fishing") ~ "Agriculture (including fishing)")) 


a <- x |> 
  dplyr::filter(year == 2022) |> 
  dplyr::group_by(ag_fish) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::filter(!is.na(ag_fish)) |> 
  dplyr::rename(sector = ag_fish)

b <- x |> 
  dplyr::filter(year == 2022) |> 
  dplyr::group_by(sector) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::filter(sector %in% c("Manufacturing", "Wholesale", "Retail", "Catering"))


c <- x |> 
  dplyr::filter(year == 2022) |> 
  dplyr::group_by(food_chain) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::filter(!is.na(food_chain)) |> 
  dplyr::rename(sector = food_chain)

d <- x |> 
  dplyr::filter(year == 2022) |> 
  dplyr::group_by(total) |> 
  dplyr::summarise(value = sum(value)) |> 
  dplyr::filter(!is.na(total)) |> 
  dplyr::rename(sector = total)

fsp_table <- dplyr::bind_rows(a, b, c, d) |> 
  dplyr::mutate(sector = dplyr::case_when(sector == "Manufacturing" ~ "Food and Drink Manufacturing",
                                          sector == "Wholesale" ~ "Food and Drink Wholesaling",
                                          sector == "Retail" ~ "Food and Drink Retailing",
                                          sector == "Catering" ~ "Food and Drink Non-Residential Catering",
                                          .default = sector),
                value = value/1000) |> 
  dplyr::select(Sector = sector, `Employees (millions)` = value)

fsp_table |> knitr::kable(digits = 3,format = "pipe")

#auk_table

```

